Extended Data

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
italicize variables & avoid latex
-Italic font used for variables, genes and non-English languages; bold font for vectors
- Equations and special characters in Word in ‘normal text’ or Symbol font (Word Equation Editor avoided where possible)

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree stem growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021;@cabon_crossbiome_2022].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

<!--![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ $^\circ$C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }-->

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (CTW). 
CTW was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (CTW *T~max~*) used as our independent variable.

<!--![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (CTW), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }-->

For ring- and diffuse- porous species at both sites, warmer CTW *T~max~* resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with CTW *T~max~* (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ $^\circ$C for ring porous species and 2.8-2.9 days/ $^\circ$C for diffuse porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring porous species and 0.9-4.2 days/ $^\circ$C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the CTW *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig. `r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig. `r schematic_summary`, `r EDF_pheno_all`).

In contrast to the pronounced effects of CTW *T~max~* on the timing of growth, its effects on *g~max~* and $\Delta DBH$ were inconsistent, and usually weak (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to CTW *T~max~* for ring- and diffuse- porous species.
$\Delta DBH$ displayed a weak positive relationship (+0.003-0.008 cm/ $^\circ$C) with CTW *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ $^\circ$C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).

@@

To understand how annual radial stem growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35$^\circ$N latitude) to Michigan (45.56$^\circ$N latitude) and spanning a 15 $^\circ$C range in April *T~max~*.
Across all chronologies, the standardized ring-width index (RWI) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`). 
There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
In contrast, RWI was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and RWI (`r EDF_tree_rings_TMIN`).

To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between RWI and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of RWI to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.

To test whether there may be an enhancement of growth by warmer spring temperatures that was offset by the negative effect of high summer temperatures, we tested for the joint effects of April and June-July *T~max~* on RWI.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (p $\le$ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).

<!--![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**. Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (RWI) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`). Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }-->

@@

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem diameter growth but have little effect on annual increments (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and NEE[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase ANPP~woody~, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014;@cabon_crossbiome_2022], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].

<!--discussion of extra C-->
Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
It remains theoretically possible that warm springs could augment ANPP~woody~, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or ANPP~woody~.

The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when GPP of a mature forest was increased through experimental CO~2~ enrichment, ANPP~woody~ remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
Thus, observed augmentation of NEE by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.

>>>>

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[<!--@jeong_phenology_2011;--><!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@keenan_net_2014;@finzi_carbon_2020;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (_GPP_) and net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012;@buermann_widespread_2018].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (*ANPP~woody~*) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of *ANPP~woody~* to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree stem growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and *ANPP~woody~* respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[<!--@jeong_phenology_2011;-->@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], _GPP_[@delpierre_wood_2016], or _NEE_[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014;@zani_increased_2020], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zohner_how_2021;@cabon_crossbiome_2022].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height (*∆DBH*; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

<!--![**Figure 1 | Summary of temperate deciduous tree growth responses to warmer spring temperatures.** (a) Schematic illustrating parameters of interest and summarizing predominant responses of each to warmer maximum temperatures during a 'critical temperature window', defined as the period with the strongest temperature control over *DOY~25~*; (b) Variable definitions and summary of responses to warmer spring temperatures at two temperate forests -- Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest -- and for two groups of broadleaf deciduous species (RP=ring porous; DP=diffuse porous), where up and down arrows indicate increases and decreases that are significant both statistically (p<0.05) and biologically (effect size >3%/ °C), '-' indicates no significant correlation, and 'mixed' indicates a mix of significant and non-significant correlations, often in different directions.](tables_figures/schematic_summary_with_table.png){ width=70% }-->

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (_CTW_). 
_CTW_ was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The _CTW_ was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (_CTW_ *T~max~*) used as our independent variable.

<!--![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each ForestGEO plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring porous and diffuse porous trees at SCBI (2011-2020) and Harvard Forest (1999-2003). Mean *DOY~25~*, *DOY~50~*, and *DOY~75~* were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean maximum temperature (*T~max~*) was calculated for each xylem architecture/site combination over the respective critical temperature window (_CTW_), then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to the _CTW_ *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest).](tables_figures/DOYtiming_allyears.png){ width=25% }-->

For ring- and diffuse- porous species at both sites, warmer _CTW_ *T~max~* resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies<!--[@parmesan_globally_2003;--><!--@ibanez_forecasting_2010;--><!--@friedl_tale_2014]-->, leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.5 days/°C at SCBI (p=0.016) and 2.4 days/ °C at Harvard Forest (p=0.1).
Similarly, with the exception of ring porous *DOY~75~* at SCBI, the stem growth milestones *DOY~25~*, *DOY~50~*, and *DOY~75~* decreased with _CTW_ *T~max~* (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, *DOY~25~*, *DOY~50~*, and *DOY~75~* advanced 0-1.7 days/ °C for ring porous species and 2.8-2.9 days/ °C for diffuse porous species at SCBI, and 10.3-12.3 days/ °C for ring porous species and 0.9-4.2 days/°C for diffuse porous species at Harvard Forest (`r EDT_phenology_parameters`).

Whereas the length of time between canopy greenup and senescence (*i.e.,* the day when greenness dropped below 90% of its peak) increased with the _CTW_ *T~max~* of the porosity group containing the dominant canopy species at each site (diffuse porous at SCBI, ring porous at Harvard Forest; Fig. `r DOY_timing`a-b), there was no consistent lengthening of *L~pgs~* (Fig. `r schematic_summary`, `r EDF_pheno_all`).

In contrast to the pronounced effects of _CTW_ *T~max~* on the timing of growth, its effects on *g~max~* and *∆DBH* were inconsistent, and usually weak (Fig. `r schematic_summary`, `r EDF_pheno_all`). 
Specifically, *g~max~*, which occurred on average within 5 days of *DOY~50~*, displayed extremely small negative changes (Harvard Forest) or changes in opposite directions (SCBI) in relationship to _CTW_ *T~max~* for ring- and diffuse- porous species.
*∆DBH* displayed a weak positive relationship (+0.003-0.008 cm/ °C) with _CTW_ *T~max~* at SCBI and a weak negative relationship (-0.02-0.003 cm/ °C) at Harvard Forest (`r EDF_pheno_Tmax_climwin`).

@@

To understand how annual radial stem growth increments have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies of `r n_chronologies_SCBI` species at SCBI[@helcoski_growing_2019] and `r n_chronologies_HF` species at Harvard Forest (`r EDT_species`), plus an additional `r n_chronologies-n_chronologies_SCBI-n_chronologies_HF` chronologies from `r n_cores_sites-2` sites (Fig. `r tree_rings`; `r EDF_map`; `r EDT_tree_ring_summary`).
In total, our analysis included `r n_chronologies` chronologies representing `r n_cores_species` broadleaf species at `r n_cores_sites` sites distributed from Alabama (34.35°N latitude) to Michigan (45.56°N latitude) and spanning a 15 °C range in April *T~max~*.
Across all chronologies, the standardized ring-width index (_RWI_) was significantly (95% confidence interval did not include 0) positively correlated with April *T~max~* for only 2% of chronologies: 1 of 142 ring porous and 4 of 66 diffuse porous species-site combinations (`r EDT_tree_ring_summary`). 
There were even fewer significant positive correlations with March and May *T~max~*: 0% and `r round(2/(142+66),2)*100`%, respectively (`r EDT_tree_ring_summary`).
In contrast, _RWI_ was frequently negatively correlated with *T~max~* during peak growing season months (May-August), with significant correlations for 52% (May: 45/141, Jun: 107/141, Jul: 91/141, Aug: 53/141) and 46% (May: 10/66, Jun: 52/66, Jul: 36/66, Aug: 23/66) of species-site-month combinations for ring- and diffuse-porous species, respectively.
*T~min~* generally exhibited weaker relationships to annual growth than *T~max~*, with few significant correlations between spring *T~min~* and _RWI_ (`r EDF_tree_rings_TMIN`).

To test whether warm springs might result in storage of non-structural carbohydrates (NSC) that would augment growth the following year[@richardson_seasonal_2013], we extended the analysis to examine correlations between _RWI_ and *T~max~* in the previous year (`r EDF_tree_rings_TMAX_extended`).
This revealed little effect of previous spring temperatures on annual growth, with significant positive correlations of _RWI_ to previous March or April *T~max~* for 5/142 ring porous chronologies and to previous April or May *T~max~* for 7/66 diffuse porous chronologies.

To test whether there may be an enhancement of growth by warmer spring temperatures that was offset by the negative effect of high summer temperatures, we tested for the joint effects of April and June-July *T~max~* on _RWI_.
Results were qualitatively similar to the univariate correlations (Fig. `r tree_rings`), with significant (p ≤ 0.05) positive correlations to April *T~max~* for only 4% of chronologies and significant negative correlations with June-July *T~max~* for 77% of chronologies (`r EDT_tree_ring_summary`).

<!--![**Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly mean maximum temperatures (*T~max~*), for `r n_chronologies` chronologies from `r n_cores_sites` sites across eastern North America**. Colors indicate the correlation between monthly *T~max~* and a dimensionless ring width index (_RWI_) derived from the multiple trees that form each chronology and emphasizing interannual variability associated with climate. Chronologies are grouped by xylem porosity and ordered by mean April *T~max~*. Plots are annotated to highlight records from our two focal sites, the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest (HF; `r EDT_species`). Sites included are mapped in `r EDF_map `, species analyzed and numbers of significant correlations to *T~max~* are summarized in `r EDT_tree_ring_summary`, and chronology details are given in `r chron_table`.](tables_figures/tree_rings_tmax.png){ width=100% }-->

@@

<!--summary, setting expectation of extra C-->
Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem diameter growth but have little effect on annual increments (Figs. `r schematic_summary`-`r tree_rings`). 
The observed advance in stem growth under warmer springs parallels advances observed for canopy foliage phenology (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020<!--;@stuble_plant_2021-->] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase *ANPP~woody~*, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. `r tree_rings`).
Our observations suggest that the deceleration of stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is driven by cues other than photosynthate limitation, such as water stress[@delpierre_wood_2016;@zweifel_why_2021;@etzold_number_2022], nutrient limitation[@elmore_earlier_2016], daylength[@etzold_number_2022], or sink saturation[@delpierre_wood_2016;@zani_increased_2020].
This adds to a growing body of evidence for a sink limitation of stem growth[@delpierre_wood_2016;@fatichi_moving_2014;@cabon_crossbiome_2022], wherein global change factors known to enhance photosynthesis, such as longer growing seasons or elevated CO~2~, do not cause a corresponding increase in stem growth[@etzold_number_2022;@jiang_fate_2020].

<!--discussion of extra C-->
Combined with widespread observations that warming springs tend to lengthen the season of CO~2~ uptake[@lu_no_2022;@finzi_carbon_2020] and increase net annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014;<!--@crabbe_extreme_2016;-->@zhou_explaining_2016;@fu_maximum_2019;@finzi_carbon_2020], our findings imply a lengthening of the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem and an increase in C allocated to functions other than stem expansion in the current or following year.
It remains theoretically possible that warm springs could augment *ANPP~woody~*, which, although routinely calculated based on stem growth, can be partially decoupled from it through differences in wood density or C content[@delpierre_wood_2016].
Extra C fixed in years with warm springs could potentially be allocated to formation of more C-dense wood, either through enhanced cell wall thickening (a process that lags behind stem expansion[@cuny_woody_2015]) or to a higher ratio of high-density latewood to lower-density earlywood.
However, existing evidence indicates that vessel features are most strongly controlled by summer drought stress in the previous (earlywood) or current (latewood) year, while warm springs have a neutral or negative effect on latewood width[@tardif_influence_2006;@roibu_climatic_2020;@kern_multiple_2013]<!--tardif is essential here-->.
Thus, it is unlikely that warm springs have a positive effect on total C content of annual rings or *ANPP~woody~*.

The fate of any additional C fixed during years with warm springs remains unresolved, but possible destinations -- including respiration, NSC storage, and production of foliage, reproductive structures, roots[@zohner_how_2021], or root exudates -- generally have shorter residence times than woody growth.
Indeed, when _GPP_ of a mature forest was increased through experimental CO~2~ enrichment, *ANPP~woody~* remained unchanged, while additional C was released back to the atmosphere on relatively short time scales through enhanced respiration[@jiang_fate_2020].
Consistent with this, it has been observed that C gains from an earlier spring can be offset through fall or winter respiration[@oishi_warmer_2018], although even the C in shorter-lived pools would often be carried over into the following year[@trumbore_quantifying_2002].
Thus, observed augmentation of _NEE_ by warm springs[@zhou_explaining_2016;@fu_maximum_2019;@keenan_net_2014] is likely to be compensated by increased respiration in subsequent years.
